Unrecognized States: A Theory of Self-Determination and Foreign Influence

### 2016-07-15 draft.tex
change to 1.25 margins
switch to endnotes
make reference citations in text conform to JLEO style sheet

\documentclass[11pt,letterpaper, notitlepage]{article}
\usepackage{epsf,graphicx,psfrag}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage[reqno]{amsmath}
\usepackage{endnotes}
\usepackage{natbib}
\usepackage{amsfonts}
\usepackage{amssymb,enumerate}
\usepackage{dcolumn}
\usepackage{verbatim}
%\usepackage{hypertex}
\usepackage[pdftex, bookmarks, colorlinks=true, plainpages=false, citecolor = blue, urlcolor = blue, filecolor = blue, breaklinks]{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[hmargin=3.0cm,vmargin=3.0cm]{geometry}
%\usepackage{geometry}
\usepackage{nicefrac}
\usepackage{pdfsync}
\usepackage{url,bm}
\usepackage{pst-all}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage[T1]{fontenc}
\usepackage[sc]{mathpazo}
\usepackage{graphicx,verbatim,scalefnt}
\pdfoptionpdfminorversion=6 %gets rid of warning about pdf version

@@

The costs to the secessionists are arguably the highest. Non-recognition locks unrecognized states out of a states-only club whose members enjoy benefits in terms of both security and economic integration. Unrecognized states do not benefit from the norm of territorial integrity that reduces the cost of territorial defense by increasing the chance that foreign powers will intervene against, or at least sanction, an invader.\footnote{See Zacher (2001) on the norm of territorial integrity.} Non-recognition also prevents entry into multilateral and bilateral trade agreements, dramatically reduces access to foreign aid, and excludes their territory from international legal frameworks, making it harder to secure foreign investment (Milhalkanian 2004; Caspersen 2013: 40-45).\footnote{Transnistria represents a case where the gains from smuggling, particularly for the elite, may have outweighed these other economic costs in the 1990s (King 2001). However, as Moldova deepens ties with the EU, Transnistria's lack of access to EU markets and EU aid becomes a much larger relative cost to bear (Tudoroiu 2011).} %In their most direct form -- loans and foreign aid -- these benefits are referred to as ``rents to sovereignty" (Collier and Hoeffler 2005).

@@

The first literature to address unrecognized states directly was grounded in comparative politics, and a robust area-studies literature exists around each of the current cases of unrecognized statehood.\footnote{Lynch (2004), King (2001), Stanislawski (2008), and Bakke et al. (2013) provide notable treatments of the Former Soviet cases, and a pair of edited volumes by Bahcheli, Bartmann, and Srebrnik (2004) and Kingston and Spears (2004) each compile broader sets of case studies.} More recent literature has addressed wider ranges of cases and made important conceptual progress identifying patterns and commonalities across cases (e.g. Pegg 1998; Kolst\o \ 2006; Geldenhuys 2009; Caspersen and Stansfield 2011; Caspersen 2013). Bakke (2011) summarizes four explanations for the persistence of unrecognized states, each of which plays a role in the theory we develop here. First, economic and political elites benefit from the status quo (e.g. King 2001). Second, the secessionists have sufficient military capacity to make military reconquest costly (e.g. Kolst\o \ 2006). Third, secessionists enjoy external support from a patron (e.g. Lynch 2002; Stanislawski 2008). Lastly, the governments of unrecognized states are internally legitimate, enjoying support from residents of the unrecognized state (e.g. Lynch 2002, Kolst\o \ 2006).

@@

Although we are not aware of any formal models of unrecognized statehood there is a growing economics literature that employs game theoretic models of conflict. Grossman (2004) offers an early model of territorial disputes in which the divisibility of the contested territory is an important parameter. In Hirshleifer, Boldrin and Levin (2009), conflict can be avoided when there is common knowledge, a common rate of time preference, and players can make a series of small concessions. In our context, the contested territory is indivisible; this implies that peace would require very large concessions so that this pathway of using small compromises to avoid conflict is not available.

@@

Another strand of the literature explicitly models the bargaining process between potential adversaries, either with asymmetric (Yared 2010) or complete information (Jackson and Morelli 2007, Schwartz and Sonin 2008, Bevi\'{a} and Corch\'{o}n 2010, McBride and Skaperdas 2014). Acemoglu, Golosov, Tsyvinski and Yared (2012) develop a dynamic model of resource wars featuring limited commitment. We abstract from questions of resource allocation and bargaining in order to focus on the dynamics created by the involvement of outside players: in the case of unrecognized states, the international community and patron states.

@@

One innovation of our model is the introduction of international actors. While the role of outside actors in determining the duration and outcome of civil conflicts is well documented (e.g., Elbadawi and Sambanis 2000; Regan 2002; Balch-Lindsay, Enterline, and Joyce 2008), the role of these actors has not been incorporated into the modeling of these conflicts.\footnote{One exception is van Houten (1998), who models the patron state ("reference state'') as a player in ethnic conflicts but otherwise takes an approach quite different from ours.} This is true even of work that addresses the role of outside actors as potential third-party enforcers (Walter 1997, 2002).

@@

Player $p$ most prefers recognized independence, aligning its interests with the secessionists. We refer to $p$ as the patron because $p$ contributes resources to the unrecognized state in the status quo equilibrium. Patrons choose to contribute resources to secessionists for one or more of several reasons: 1) As an efficient mechanism for imposing costs on the home state (Salehyan, Gleditsch and Cunningham 2012), e.g. as Russia does to Georgia via South Ossetia and Abkhazia; 2) ethnic solidarity with the secessionists (e.g. Turkey's support of the Turkish Republic of Northern Cyprus); 3) hope of eventual annexation of the disputed territory (e.g. Armenia's support of Nagorno-Karabakh).\footnote{While annexation is appealing to many patrons (and some unrecognized states), annexation is not an outcome we model. International norms against irredentism are strong, and the costs of annexing an unrecognized state appear to be prohibitively large in most cases (e.g. Zacher 2001). Prior to Russia's annexation of Crimea, all post-WWII cases of annexation involved the annexation of colonial territory, rather than parts of the metropole. Israel's occupation of the Golan Heights and the West Bank are possible exceptions here.}

@@

\subsubsection{The Low Payoffs from Ceding}
Aside from the central assumption that the government of the home state most prefers reunification and the secessionists most prefer independence accompanied by international recognition, we will for the most part leave the model general enough to incorporate a range of preferences and capabilities for players $g$ and $s$. %Our results will specify parameter ranges for these preferences over which outcomes of interest---such as perpetual unrecognized statehood---can be supported in equilibrium.
We do, however, assert that the payoffs for the party that cedes the issue of status (independence vs. reunification) are consistently low. This reflects a combination of two factors. First, the issue of status is indivisible\footnote{Either the unrecognized state has sovereignty over its territory and is co-equal with the home state, or the secessionist region (and its government) are subordinate to the central government.} and highly valued by each side, making its surrender undesirable. Various forms of ethno-nationalism often motivate secession, and the values attached to independence by secessionists (and to reunification by citizens of the home state) are generally large relative to the values placed on economic prosperity and other goals. Second, many of the payments that could be offered are not credible (e.g. Schultz 2010).

The difficulty of making credible payments in exchange for status is one clearly demonstrated in the civil war literature (e.g. Licklider 1995; Walter 1997, 2002; Fearon and Laitin 2007; Doyle and Sambanis 2006). Unrecognized states generally constitute ``sons of the soil" conflicts in which the central government cannot credibly commit to preserving the local demographic and political dominance of the secessionist elite once the disputed territory reverts to central government control (Weimer 1978; Fearon 2004). While the central government might initially grant the secessionist elite a high level of autonomy in exchange for agreeing to reunification, the level of autonomy is likely to decrease over time, perhaps quite quickly. The payments that can be offered by the unrecognized state to the central government are similarly small or unenforceable: once recognition is achieved, the secessionists have little reason to uphold any prior commitments.

@@

\subsection{If the Patron Withdraws Support}
Support by a foreign patron is, in almost all cases, necessary for the persistence of unrecognized statehood (Kolst\o \ 2006; Caspersen 2013). When there is no patron, or when the patron withdraws its support, military reconquest by the home state is likely. The only unrecognized state currently in existence which has been able to survive without a patron is Somaliland. Somaliland has been able to survive as long as it has because of the extreme weakness of the home state (Somalia).\footnote{Owtram (2011) argues that, over time, unrecognized states can become sufficiently "consolidated" as to no longer be dependent on a patron for support. However, no cases other than Somaliland support this claim empirically.}

@@

As noted in the section on outside interactions between the patron and player $c$, the patron's decision to withdraw support for the secessionists is sometimes motivated by interactions between the patron and $c$ that we do not model directly. Empirically, we observe a number of cases in which $c$ applies pressure directly to encourage the patron to withdraw support from the unrecognized state. In an extreme example involving both sanctions and direct military confrontation, NATO coerced Serbia into, among other things, withdrawing its support from Republika Srpska and Republika Srpska Krajina, both of which were then reconquered militarily.\footnote{For an excellent discussion of the case of Republika Srpska, see Zahar 2004.} %In the final section of the paper, we consider direct coercion of the patron among the strategies available to player $c$.

@@

Serving as a third-party guarantor of autonomy rights is a way for player $c$ to potentially overcome problems of indivisibility and commitment and help the parties reach a credible compromise on status (Walter 2002). However, this strategy is only tenable when the only impediment to settlement is the unenforceability of a bargain, and when $c$ is credible as an enforcer of that bargain.

@@

>>>>

\documentclass[11pt,letterpaper, notitlepage]{article}
\usepackage{epsf,graphicx,psfrag}
\usepackage{natbib}
\usepackage{setspace}
\usepackage{fullpage}
\usepackage[reqno]{amsmath}
\usepackage{endnotes}
\usepackage{natbib}
\usepackage{amsfonts}
\usepackage{amssymb,enumerate}
\usepackage{dcolumn}
\usepackage{verbatim}
%\usepackage{hypertex}
\usepackage[pdftex, bookmarks, colorlinks=true, plainpages=false, citecolor = blue, urlcolor = blue, filecolor = blue, breaklinks]{hyperref}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage[hmargin=3.25cm,vmargin=3.0cm]{geometry}
%\usepackage{geometry}
\usepackage{nicefrac}
\usepackage{pdfsync}
\usepackage{url,bm}
\usepackage{pst-all}
\usepackage{pdflscape}
\usepackage{rotating}
\usepackage[T1]{fontenc}
\usepackage[sc]{mathpazo}
\usepackage{graphicx,verbatim,scalefnt}
\pdfoptionpdfminorversion=6 %gets rid of warning about pdf version

\let\footnote=\endnote

@@

The costs to the secessionists are arguably the highest. Non-recognition locks unrecognized states out of a states-only club whose members enjoy benefits in terms of both security and economic integration. Unrecognized states do not benefit from the norm of territorial integrity that reduces the cost of territorial defense by increasing the chance that foreign powers will intervene against, or at least sanction, an invader.\footnote{See Zacher (2001) on the norm of territorial integrity.} Non-recognition also prevents entry into multilateral and bilateral trade agreements, dramatically reduces access to foreign aid, and excludes their territory from international legal frameworks, making it harder to secure foreign investment (Milhalkanian, 2004; Caspersen, 2013: 40-45).\footnote{Transnistria represents a case where the gains from smuggling, particularly for the elite, may have outweighed these other economic costs in the 1990s (King, 2001). However, as Moldova deepens ties with the EU, Transnistria's lack of access to EU markets and EU aid becomes a much larger relative cost to bear (Tudoroiu, 2011).} %In their most direct form -- loans and foreign aid -- these benefits are referred to as ``rents to sovereignty" (Collier and Hoeffler 2005).

@@

The first literature to address unrecognized states directly was grounded in comparative politics, and a robust area-studies literature exists around each of the current cases of unrecognized statehood.\footnote{Lynch (2004), King (2001), Stanislawski (2008), and Bakke et al. (2013) provide notable treatments of the Former Soviet cases, and a pair of edited volumes by Bahcheli et al. (2004) and Kingston and Spears (2004) each compile broader sets of case studies.} More recent literature has addressed wider ranges of cases and made important conceptual progress identifying patterns and commonalities across cases (e.g. Pegg, 1998; Kolst\o, \ 2006; Geldenhuys, 2009; Caspersen and Stansfield, 2011; Caspersen 2013). Bakke (2011) summarizes four explanations for the persistence of unrecognized states, each of which plays a role in the theory we develop here. First, economic and political elites benefit from the status quo (e.g. King, 2001). Second, the secessionists have sufficient military capacity to make military reconquest costly (e.g. Kolst\o, \ 2006). Third, secessionists enjoy external support from a patron (e.g. Lynch, 2002; Stanislawski, 2008). Lastly, the governments of unrecognized states are internally legitimate, enjoying support from residents of the unrecognized state (e.g. Lynch, 2002, Kolst\o, \ 2006).

@@

Although we are not aware of any formal models of unrecognized statehood there is a growing economics literature that employs game theoretic models of conflict. Grossman (2004) offers an early model of territorial disputes in which the divisibility of the contested territory is an important parameter. In Hirshleifer et al., (2009), conflict can be avoided when there is common knowledge, a common rate of time preference, and players can make a series of small concessions. In our context, the contested territory is indivisible; this implies that peace would require very large concessions so that this pathway of using small compromises to avoid conflict is not available.

@@

Another strand of the literature explicitly models the bargaining process between potential adversaries, either with asymmetric (Yared, 2010) or complete information (Jackson and Morelli, 2007; Schwartz and Sonin, 2008; Bevi\'{a} and Corch\'{o}n, 2010; McBride and Skaperdas, 2014). Acemoglu et al. (2012) develop a dynamic model of resource wars featuring limited commitment. We abstract from questions of resource allocation and bargaining in order to focus on the dynamics created by the involvement of outside players: in the case of unrecognized states, the international community and patron states.

@@

One innovation of our model is the introduction of international actors. While the role of outside actors in determining the duration and outcome of civil conflicts is well documented (e.g., Elbadawi and Sambanis, 2000; Regan 2002; Balch-Lindsay et al., 2008), the role of these actors has not been incorporated into the modeling of these conflicts.\footnote{One exception is van Houten (1998), who models the patron state ("reference state'') as a player in ethnic conflicts but otherwise takes an approach quite different from ours.} This is true even of work that addresses the role of outside actors as potential third-party enforcers (Walter, 1997; 2002).

@@

Player $p$ most prefers recognized independence, aligning its interests with the secessionists. We refer to $p$ as the patron because $p$ contributes resources to the unrecognized state in the status quo equilibrium. Patrons choose to contribute resources to secessionists for one or more of several reasons: 1) As an efficient mechanism for imposing costs on the home state (Salehyan et al., 2012), e.g. as Russia does to Georgia via South Ossetia and Abkhazia; 2) ethnic solidarity with the secessionists (e.g. Turkey's support of the Turkish Republic of Northern Cyprus); 3) hope of eventual annexation of the disputed territory (e.g. Armenia's support of Nagorno-Karabakh).\footnote{While annexation is appealing to many patrons (and some unrecognized states), annexation is not an outcome we model. International norms against irredentism are strong, and the costs of annexing an unrecognized state appear to be prohibitively large in most cases (e.g. Zacher, 2001). Prior to Russia's annexation of Crimea, all post-WWII cases of annexation involved the annexation of colonial territory, rather than parts of the metropole. Israel's occupation of the Golan Heights and the West Bank are possible exceptions here.}

@@

\subsubsection{The Low Payoffs from Ceding}
Aside from the central assumption that the government of the home state most prefers reunification and the secessionists most prefer independence accompanied by international recognition, we will for the most part leave the model general enough to incorporate a range of preferences and capabilities for players $g$ and $s$. %Our results will specify parameter ranges for these preferences over which outcomes of interest---such as perpetual unrecognized statehood---can be supported in equilibrium.
We do, however, assert that the payoffs for the party that cedes the issue of status (independence vs. reunification) are consistently low. This reflects a combination of two factors. First, the issue of status is indivisible\footnote{Either the unrecognized state has sovereignty over its territory and is co-equal with the home state, or the secessionist region (and its government) are subordinate to the central government.} and highly valued by each side, making its surrender undesirable. Various forms of ethno-nationalism often motivate secession, and the values attached to independence by secessionists (and to reunification by citizens of the home state) are generally large relative to the values placed on economic prosperity and other goals. Second, many of the payments that could be offered are not credible (e.g. Schultz, 2010).

The difficulty of making credible payments in exchange for status is one clearly demonstrated in the civil war literature (e.g. Licklider, 1995; Walter 1997, 2002; Doyle and Sambanis, 2006; Fearon and Laitin, 2007). Unrecognized states generally constitute ``sons of the soil" conflicts in which the central government cannot credibly commit to preserving the local demographic and political dominance of the secessionist elite once the disputed territory reverts to central government control (Weimer, 1978; Fearon, 2004). While the central government might initially grant the secessionist elite a high level of autonomy in exchange for agreeing to reunification, the level of autonomy is likely to decrease over time, perhaps quite quickly. The payments that can be offered by the unrecognized state to the central government are similarly small or unenforceable: once recognition is achieved, the secessionists have little reason to uphold any prior commitments.

@@

\subsection{If the Patron Withdraws Support}
Support by a foreign patron is, in almost all cases, necessary for the persistence of unrecognized statehood (Kolst\o, \ 2006; Caspersen, 2013). When there is no patron, or when the patron withdraws its support, military reconquest by the home state is likely. The only unrecognized state currently in existence which has been able to survive without a patron is Somaliland. Somaliland has been able to survive as long as it has because of the extreme weakness of the home state (Somalia).\footnote{Owtram (2011) argues that, over time, unrecognized states can become sufficiently "consolidated" as to no longer be dependent on a patron for support. However, no cases other than Somaliland support this claim empirically.}

@@

As noted in the section on outside interactions between the patron and player $c$, the patron's decision to withdraw support for the secessionists is sometimes motivated by interactions between the patron and $c$ that we do not model directly. Empirically, we observe a number of cases in which $c$ applies pressure directly to encourage the patron to withdraw support from the unrecognized state. In an extreme example involving both sanctions and direct military confrontation, NATO coerced Serbia into, among other things, withdrawing its support from Republika Srpska and Republika Srpska Krajina, both of which were then reconquered militarily.\footnote{For an excellent discussion of the case of Republika Srpska, see Zahar, 2004.} %In the final section of the paper, we consider direct coercion of the patron among the strategies available to player $c$.

@@

Serving as a third-party guarantor of autonomy rights is a way for player $c$ to potentially overcome problems of indivisibility and commitment and help the parties reach a credible compromise on status (Walter, 2002). However, this strategy is only tenable when the only impediment to settlement is the unenforceability of a bargain, and when $c$ is credible as an enforcer of that bargain.

@@

\newpage

\theendnotes
